TacticToe: Learning to Prove with Tactics

### 2018-03-19 tactictoe_jv.tex
Plan and diagram in the introduction

\usepackage{fancyvrb}
\usepackage{tikz} % graphics
% graphs
\usetikzlibrary{arrows,shapes}
\usetikzlibrary{trees,positioning,fit}
\tikzstyle{block} = [rectangle, draw,% fill=blue!20,
    text width=5em, text centered, rounded corners, minimum height=4em]
\tikzstyle{line}=[draw]
\tikzstyle{cloud} = [draw, ellipse,%fill=red!20,
    node distance=3cm,
    minimum height=2em]

@@

\section{Introduction}
Many of the state-of-the-art interactive theorem provers (ITPs) such as
  \holfour~\cite{hol4}, \hollight~\cite{Harrison09hollight},
  \isabelle~\cite{isabelle}
  and \coq~\cite{coq-book} provide high-level parameterizable tactics for constructing proofs.
  Tactics analyze the current proof state (goal and
  assumptions) and apply non-trivial proof transformations.
  Formalized proofs take advantage of different levels of automation:
  specialized rules, theory-based strategies and general purpose strategies.
  Thanks to progress in proof
  automation, developers can delegate more and more complicated proof
  obligations to general purpose strategies. Those are implemented by automated
  theorem provers (ATPs) such as \eprover~\cite{eprover}. Communication
  between
  an ITP and ATPs is made possible by a ``hammer''
  system~\cite{hammers4qed,tgck-cpp15}. It acts as an interface by performing
  premise selection, translation and proof reconstruction.
  Yet, ATPs are far from foolproof and more precise user-guidance, achieved
  by applying a particular sequence of specialized rules, is almost always
  necessary to develop a mathematical theory.

  In this work, we develop in the ITP \holfour a procedure to select suitable
  tactics depending on the current proof state by learning
  from previous proofs. Backed by this machine-learned guidance, our prover
  \tactictoe relies on \emph{Monte
  Carlo tree search}~\cite{montecarlo} to find sequences of tactic applications
  leading to an ITP proof.
  To this end, we \emph{extract and record} tactic invocations from the ITP
  proofs (Section~\ref{sec:recording}) and
  \emph{build efficient machine learning classifiers} based on such training
  examples (Section~\ref{s:prediction}).  The learned data serves as
  guidance for the Monte Carlo tree search that explores different
  proof paths (Section~\ref{sec:proofsearch}). The result, if
  successful, is a verified human-level proof composed of \holfour
  tactics (Section~\ref{sec:proofdisplay}). We evaluate the system on a large
  set of theorems originating
  from \holfour (Section~\ref{s:experiments}).

\subsection{The problem}
% How a user works with an ITP and how this gives rise to a machine learning problem.
\todo{RK: this is just a first draft -- feel free to edit lots!}
An ITP is a development environment for the construction of formal proofs in which it is possible, but not common, to write a proof as a sequence of applications of primitive inference rules.
The common approach, however, is to use and devise high-level tools and automation that abstract over useful ways of producing proof pieces.
A specific, prevalent instance of this approach is the use of \emph{tactics} for \emph{backward} or goal-directed proof.
Here, the ITP user operates on a proof state, initially the desired conjecture, and applies tactics that transform the proof state until it is solved (or until giving up).
Each tactic performs a sound transformation of the proof state: essentially, it promises that if the resulting proof state can be solved, then so can the starting one.
This gives rise to a machine learning problem: can we learn a mapping from a given proof state to the next tactic (or sequence of tactics) that will productively advance the proof towards a solution?

@@

@@

%A similar trade-off between exploration and exploitation exists in algorithm
%like Monte Carlo. However, Monte Carlo tree search metrics are based on many
%simulations and do not reflect our exploration strategy that only
\paragraph{Node selection}
Through node extension steps the search tree grows and the number of paths to
explore increases. To decide which node to extend next, an evolving value for
each node is given in Definition~\ref{def:value}. The algorithm that performs
node selection starts from the root of the search tree. From the current node,
the function \texttt{CompareChildren} chooses among the children of its open
goal the one with the highest value.

@@

\section{Feature Extraction}\label{sec:features}
%We explain the construction of the similarity measure by which relevant
%objects are predicted

@@

\subsection{Similarity}\label{sec:predictions}
We estimate the similarity (or co-distance) between two objects $o_1$ and $o_2$
through their respective feature sets $f_1$ and $f_2$.
The estimation is based on the frequency of the features in the intersection of
$f_1$ and $f_2$. A good rule of thumb is that the rarer the shared features
are, the more similar two goals should be. The relative rarity of each feature
can be estimated by calculating its TF-IDF weight~\cite{Jones72astatistical}.
We additionally raise these weighs to the sixth power giving
even more importance
to rare features. This modification was found experimentally optimal in
\cite{ckju-pxtp13}.

@@

In Table~\ref{tab:tuning}, the success rate of the different \tactictoe
strategy is presented.
The first four techniques are tested relative to the same baseline indicated in
the table by the tag ``default''.  This default strategy relies on a tactic
time out of 0.05s, an orthogonalization radius of 20 and a policy coefficient of
0.5.
The two last techniques are tested relative
to a baseline consisting of the best set of parameters discovered so far and
are marked with the improvement over the ``default'' strategy.
In each subtable, each experiment differs from the current baseline by exactly
one parameter.

Thanks to the larger search time available in this experiment, the timeout for
each tactic can be increased from the 0.02 seconds used in \tactictoe's initial
experiments~\cite{tgckju-lpar17} to 0.05
seconds. Removing tactics with similar effect as performed by the
orthogonalization process is only beneficial when running \tactictoe for short
period of time. It seems that the allotted time allows a strategy
running without orthogonalization to catch up.
It is best to predict a list of 16 theorems to instantiate arguments of type
theorem list in tactics. At any rate, argument prediction for abstracted
tactics is the technique that have the highest impact on the success of
\tactictoe. Integration of ATPs being a specialization of this technique
contributes significantly as well. Experiments involving \eprover, run the ATP
using a separate process asynchronously with a timeout of 5 seconds and 128 premises.
The fact that \metis outperforms \eprover in this setting is due to the fact
that \metis is run with a very short timeout, allowing to close different part
of search tree quickly. Because \metis is weaker that \tactictoe as an ATP, it
is given less predictions. But even if an essential lemma for the proof of a
goal is not predicted, a modification on the goal performed by a tactic can
change its prediction to include the necessary lemma.
This effect minimizes the drawback of relying on a small number of predictions.

@@

The extracted proof script of $\mathfrak{T}$ is $P_{goal}(g_{root})$.
We minimally improve it by substituting \texttt{THENL} by \texttt{THEN} when the
list of goals is a singleton and removing \texttt{THENL []} during the
extraction phase.
Further post-processing such as
eliminating unnecessary tactics and theorems has been developed and
improves the user experience greatly~\cite{DBLP:conf/sefm/Adams15}.

@@

\section{Case Study}
Since the proofs generated by \tactictoe are meant to be parts of a \holfour development, it interesting to compare their quality with the original human proofs.
The quality of a proof in \holfour can be measured in terms of length, readability, maintainability and verification speed.
We study these properties in three examples taken from our full-scale experiment.
We list the time needed by \holfour to check a proof in parentheses.

We start with Example~\ref{ex:cs1}, which proves that greatest common divisors are unique.
The human formalizer recognized that it follows from two theorems.
The relevance filtering of \tactictoe is not as powerful as that used in hammers
and is therefore not able to find the \texttt{DIVIDES\_ANTISYM} property. The proof
proceeds instead by rewriting the
goal with the definitions then splitting the goal into multiple cases until the goal obligations are similar
to \texttt{DIVIDES\_ANTISYM}. As expected, \tactictoe's proof
also takes much longer to check.

>>>>

\usepackage{fancyvrb}
\usepackage{tikz} % graphics
% graphs
\usetikzlibrary{arrows,shapes}
\usetikzlibrary{trees,positioning,fit}
\tikzstyle{block} = 
  [rectangle, draw, 
   minimum width=9em, text centered, rounded corners, minimum height=3.5em]
\tikzstyle{line}=[draw]
\tikzstyle{cloud} = 
  [draw, text centered, ellipse, minimum height=3.5em, minimum width=9em]

@@

\section{Introduction}
Many of the state-of-the-art interactive theorem provers (ITPs) such as
  \holfour~\cite{hol4}, \hollight~\cite{Harrison09hollight},
  \isabelle~\cite{isabelle}
  and \coq~\cite{coq-book} provide high-level parameterizable tactics for constructing proofs.
  Tactics analyze the current proof state (goal and
  assumptions) and apply non-trivial proof transformations.
  Formalized proofs take advantage of different levels of automation:
  specialized rules, theory-based strategies and general purpose startegies.
  Thanks to progress in proof
  automation, developers can delegate more and more complicated proof
  obligations to general purpose strategies. Those are implemented by automated
  theorem provers (ATPs) such as \eprover~\cite{eprover}. Communication
  between
  an ITP and ATPs is made possible by a ``hammer''
  system~\cite{hammers4qed,tgck-cpp15}. It acts as an interface by performing
  premise selection, translation and proof reconstruction.
  Yet, ATPs are far from foolproof and more precise user-guidance, achieved
  by applying a particular sequence of specialized rules, is almost always
  necessary to develop a mathematical theory.

  In this work, we develop in the ITP \holfour a procedure to select suitable
  tactics depending on the current proof state by learning
  from previous proofs. Backed by this machine-learned guidance, our prover
  \tactictoe relies on \emph{Monte
  Carlo tree search}~\cite{montecarlo} to find sequences of tactic applications
  leading to an ITP proof.

\subsection{The problem}
An ITP is a development environment for the construction of formal proofs in which it is possible, but not common, to write a proof as a sequence of applications of primitive inference rules.
The common approach, however, is to use and devise high-level tools and automation that abstract over useful ways of producing proof pieces.
A specific, prevalent instance of this approach is the use of \emph{tactics} for \emph{backward} or goal-directed proof.
Here, the ITP user operates on a proof state, initially the desired conjecture, and applies tactics that transform the proof state until it is solved (or until giving up).
Each tactic performs a sound transformation of the proof state: essentially, it promises that if the resulting proof state can be solved, then so can the starting one.
This gives rise to a machine learning problem: can we learn a mapping from a given proof state to the next tactic (or sequence of tactics) that will productively advance the proof towards a solution?

@@

\subsection{Plan}

\begin{figure}[h]
\begin{center}
\begin{tikzpicture}[node distance = 1.9cm]
  \node [] (aa) {\textbf{Learning}};
  \node [cloud, below of=aa, node distance = 1cm] (a) {Formal library};
  \node [block, below of=a] (b) {Proof recording};
  \node [cloud, below of=b] (c) {Knowledge base}; 
  \node [block, below of=c] (e) {Training loop};
  \node [block, below of=e] (d) {Tactic predictors};

  \draw [-to,black,thick] (a) to (b);
  \draw [-to,black,thick] (b.south) to (c.north);
  \draw [-to,black,thick] (c) to [bend left]  (e);
  \draw [-to,black,thick] (e) to [bend left]  (c);
  \draw [-to,black,thick] (c.east) to [out=330,in=30] (d);

  \node [right of=aa, node distance=6.2cm] (0) {\textbf{Proving}};
  \node [cloud, below of=0, node distance = 1cm] (1) {Conjecture};
  \node [cloud, below of=1, node distance = 1.7cm] (2) {Search tree};
  \node [block, below of=2, node distance=1.3cm] (3) {Tactic policy};
  \node [above of=2, xshift=-16mm, node distance = 0.9cm] (00) {MCTS proof 
  search};
  \node [block, below of=3, node distance = 1.3cm] (4) {Tactic evaluation};
  \node [block, fit=(2) (3) (4)] (5) {};
  \node [block, below of=4, node distance = 1.7cm] (6) {Proof minimization};
  \node [cloud, below of=6, node distance = 1.6cm] (7) {Proof};
  \draw [-to,black,thick] (1) to (5);
  \draw [-to,black,thick] (5) to (6);
  \draw [-to,black,thick] (6) to (7);
  \draw [-to,black,thick] (d) to [out=0, in=180] (3);
  \draw [-to,black,thick] (d) to [out=0, in=180] (4);

\end{tikzpicture}
\end{center}
\label{fig:components}
\caption{Relation between modules of the learning and proving tool chains.
Functions are represented by rectangles and objects by ellipses.}
\end{figure}

The different components depicted in Fig~\ref{fig:components} give a 
high-level overview of the techniques necessary to generate 
tactic-level proofs by learning from recorded knowledge. From a formal library 
of ITP proofs, we extract and record tactic invocations 
in Section~\ref{sec:recording}. Based on such training examples, we
  build efficient machine learning classifiers in Section~\ref{s:prediction}. 
  At the basis of those predictions are 
  similarity measures computed from syntactic features of mathematical formulas 
  in Section~\ref{sec:features}.
  Modification of the knowledge base performed by orthogonalization 
  in Section~\ref{sec:ortho} and abstraction in Section~\ref{sec:synthesis}
  improves the quality of the databases.
  And the preselection method in Section~\ref{sec:presel} increases the speed 
  of the prediction algorithm.
  The learned predictors relying on this knowledge serves as
  guidance for the proof search algorithm in Section~\ref{sec:proofsearch} 
  which explores different proof paths in the proof tree in 
  Section~\ref{sec:prelim}. 
  The result, if
  successful, is a verified human-level proof composed of \holfour
  tactics minimized in Section~\ref{sec:proofdisplay}. And in 
  Section~\ref{s:experiments}, we evaluate the system 
  on a large set of theorems originating from \holfour.
  And generated tactic-level proofs are analysed in 
  Section~\ref{sec:case_study}.

The section order does not follow the implementation. In this way, 
we can give a description of the proof search algorithm sooner. This is 
beneficial for the reader as it shows the essential for our work before 
technical details on the prediction and recording algorithms.

@@

%A similar trade-off between exploration and exploitation exists in algorithm
%like Monte Carlo. However, Monte Carlo tree search metrics are based on many
%simulations and do not reflect our exploration strategy that only
\paragraph{Node selection}
Through node extension steps the search tree grows and the number of paths to
explore increases. To decide which node to extend next, an evolving value for
each node is given in Definition~\ref{def:value}. The algorithm that performs
node selection starts from the root of the search tree. From the current node,
the function \texttt{CompareChildren} choses among the children of its open
goal the one with the highest value.

@@

\section{Feature Extraction and Similarity Measure}\label{sec:features}
%We explain the construction of the similarity measure by which relevant
%objects are predicted

@@

We estimate the similarity (or co-distance) between two objects $o_1$ and $o_2$
through their respective feature sets $f_1$ and $f_2$.
The estimation is based on the frequency of the features in the intersection of
$f_1$ and $f_2$. A good rule of thumb is that the rarer the shared features
are, the more similar two goals should be. The relative rarity of each feature
can be estimated by calculating its TF-IDF weight~\cite{Jones72astatistical}.
We additionally raise these weighs to the sixth power giving
even more importance
to rare features. This modification was found experimentally optimal in
\cite{ckju-pxtp13}.

@@

In Table~\ref{tab:tuning}, the success rate of the different \tactictoe
strategy is presented.
The first four techniques are tested relative to the same baseline indicated in
the table by the tag ``default''.  This default strategy relies on a tactic
time out of 0.05s, an orthogonalization radius of 20 and a policy coefficent of
0.5.
The two last techniques are tested relative
to a baseline consisting of the best set of parameters discovered so far and
are marked with the improvement over the ``default'' strategy.
In each subtable, each experiment differs from the current baseline by exactly
one parameter.

Thanks to the larger search time available in this experiment, the timeout for
each tactic can be increased from the 0.02 seconds used in \tactictoe's initial
experiments~\cite{tgckju-lpar17} to 0.05
seconds. Removing tactics with similar effect as performed by the
orthogonalization process is only beneficial when running \tactictoe for short
period of time. It seems that the alloted time allows a strategy
running without orthogonalization to catch up.
It is best to predict a list of 16 theorems to instantiate arguments of type
theorem list in tactics. At any rate, argument prediction for abstracted
tactics is the technique that have the highest impact on the success of
\tactictoe. Integration of ATPs being a specialization of this technique
contributes significantly as well. Experiments involving \eprover, run the ATP
using a separate proces asynchronously with a timeout of 5 seconds and 128 premises.
The fact that \metis outperforms \eprover in this setting is due to the fact
that \metis is run with a very short timeout, allowing to close different part
of search tree quickly. Because \metis is weaker that \tactictoe as an ATP, it
is given less predictions. But even if an essential lemma for the proof of a
goal is not predicted, a modification on the goal performed by a tactic can
change its prediction to include the necessary lemma.
This effect minimizes the drawback of relying on a small number of predictions.

@@

The extracted proof script of $\mathfrak{T}$ is $P_{goal}(g_{root})$.
We minimally improve it by subtituting \texttt{THENL} by \texttt{THEN} when the
list of goals is a singleton and removing \texttt{THENL []} during the
extraction phase.
Further post-processing such as
eliminating unnecessary tactics and theorems has been developed and
improves the user experience greatly~\cite{DBLP:conf/sefm/Adams15}.

@@

\section{Case Study}\label{sec:case_study}
Since the proofs generated by \tactictoe are meant to be parts of a \holfour development, it interesting to compare their quality with the original human proofs.
The quality of a proof in \holfour can be measured in terms of length, readability, maintainability and verification speed.
We study these properties in three examples taken from our full-scale experiment.
We list the time needed by \holfour to check a proof in parentheses.

We start with Example~\ref{ex:cs1}, which proves that greatest common divisors are unique.
The human formalizer recognized that it follows from two theorems.
The relevance filtering of \tactictoe is not as powerful as that used in hammers
and is therefore not able to find the \texttt{DIVIDES\_ANTISYM} property. The proof
proceeds instead by rewriting the
goal with the definitions then spliting the goal into multiple cases until the goal obligations are similar
to \texttt{DIVIDES\_ANTISYM}. As expected, \tactictoe's proof
also takes much longer to check.
